Preliminary analysis of Greenland narwhals trajectories with smoothSDE

### 2025-01-07 R/application/CVM/fit_cvm.Rmd
R code updates

---
title: "fit_crcvm"
author: "Alexandre Delporte"
date: "2024-06-07"
output: html_document
---

@@

## Set up

```{r packages}
set.seed(42)

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
```

@@

# Set the path to the directory containing the data
par_dir=dirname(dirname(dirname(getwd()))) #parent directory 
narwhal_data_path <- file.path(par_dir,"Data", "Narwhals")

#data before exposure with first 24h removed
dataBE1=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), header = TRUE,dec = ".")

#data before exposure with first 12h removed
dataBE2=read.csv(file.path(narwhal_data_path,"DataBE2.csv"), header = TRUE,dec = ".")

@@

@@

We choose thresholds on the distance to the shore to define the ExpShore covariate.

```{r preprocess data}

D_low=0.07
D_up=3
dataBE1[dataBE1$DistanceShore> D_up,"ExpShore"]=0
dataBE1[dataBE1$DistanceShore< D_low,"ExpShore"]=1/D_low

dataBE2[dataBE2$DistanceShore> D_up,"ExpShore"]=0
dataBE2[dataBE2$DistanceShore< D_low,"ExpShore"]=1/D_low

dataAE[dataAE$DistanceShore> D_up,"ExpShore"]=0
dataAE[dataAE$DistanceShore< D_low,"ExpShore"]=1/D_low

N=length(unique(dataBE2$ID))

n_pre=length(dataBE2$time) #number of observations before exposure
n_post=length(dataAE$time) #number of observations after exposure
```

## Baseline models

We can estimate the measurement error standard deviation directly from the data. Based on accuracy studies about fastloc GPS, it should be a few tens of metres. The estimate we get from the data before exposure is 25m (std error 1m), which seems consistent with previous studies. However, when estimating this measurement error from the post exposure data, we get a higher value of 48m, which is weird : the same GPS is device is used during all the experiment, so measurement error before or after exposure should be about the same. If we fix the measurement error to 25m when estimating the response, then we can't estimate because TMB gets a non finite value of the negative log-likelihood at initial parameters. This cannot be solved by changing the other initial parameters in the optimisation. Thus we decided to fix the measurement error to different values in the baseline and the response and compare the results we get. Higher measurement error gives higher estimates for tau, whereas the estimate of nu is not affected by the changes in the measurement error.

### Baselines with constant parameters

```{r fit baseline0}

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),omega=~1)

# Fit baseline with 25m measurement error
sigma_obs=0.025
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_25m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline0_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
baseline0_100m$fit()

@@

```{r results baseline0}
estimates_bas0_25m=as.list(baseline0_25m$tmb_rep(),what="Est")
std_bas0_25m=as.list(baseline0_25m$tmb_rep(),what="Std")

estimates_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Est")
std_bas0_50m=as.list(baseline0_50m$tmb_rep(),what="Std")

estimates_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Est")
std_bas0_75m=as.list(baseline0_75m$tmb_rep(),what="Std")

estimates_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Est")
std_bas0_100m=as.list(baseline0_100m$tmb_rep(),what="Std")

@@

### Baseline with AngleNormal covariate in omega

Fixing a high standard deviation (>50m) for the measurement error tends to increase the estimate of the parameter tau. Intuitively, for a low value of the measurement error, tortuous trajectory can only be caused by less persistent movement, whereas if the measurement error is high, a part of the tortuosity can be attributed to the measurement errors, which might give higher estimate of the persistence parameter.

```{r fit baseline1}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1,1,1)

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~s(AngleNormal,k=5,bs="cs"))

# Fit baseline with 25m measurement error
sigma_obs=0.025
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_25m$update_lambda(init_lambda)

baseline1_25m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_50m$update_lambda(init_lambda)

baseline1_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_75m$update_lambda(init_lambda)

baseline1_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline1_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"),
                    map=list(log_lambda=factor(c("tau.s(ID)","nu.s(ID)",NA))))

baseline1_100m$update_lambda(init_lambda)

baseline1_100m$fit()

@@

```{r results baseline1}
estimates_bas1_25m=as.list(baseline1_25m$tmb_rep(),what="Est")
std_bas1_25m=as.list(baseline1_25m$tmb_rep(),what="Std")

estimates_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Est")
std_bas1_50m=as.list(baseline1_50m$tmb_rep(),what="Std")

estimates_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Est")
std_bas1_75m=as.list(baseline1_75m$tmb_rep(),what="Std")

estimates_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Est")
std_bas1_100m=as.list(baseline1_100m$tmb_rep(),what="Std")

plots_bas1_25m=baseline1_25m$get_all_plots(baseline=NULL,model_name="baseline1_25m",show_CI="pointwise",save=TRUE)
plots_bas1_50m=baseline1_50m$get_all_plots(baseline=NULL,model_name="baseline1_50m",show_CI="pointwise",save=TRUE)
plots_bas1_75m=baseline1_75m$get_all_plots(baseline=NULL,model_name="baseline1_75m",show_CI="pointwise",save=TRUE)
plots_bas1_100m=baseline1_100m$get_all_plots(baseline=NULL,model_name="baseline1_100m",show_CI="pointwise",save=TRUE)

@@

### Baseline with AngleNormal and DistanceShore in omega

```{r fit baseline2}

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(1/0.2^2,1/0.2^2,1,0.01,1,1)

formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                 omega=~ti(DistanceShore,k=5,bs="cs")+ti(AngleNormal,k=5,bs="cs")+ti(AngleNormal,DistanceShore,k=c(5,5),bs="cs"))

# Fit baseline with 25m measurement error
sigma_obs=0.025
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_25m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_25m$update_lambda(init_lambda)
baseline2_25m$fit()

# Fit baseline with 50m measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_50m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_50m$update_lambda(init_lambda)
baseline2_50m$fit()

# Fit baseline with 75m measurement error
sigma_obs=0.075
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_75m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_75m$update_lambda(init_lambda)
baseline2_75m$fit()

# Fit baseline with 100m measurement error
sigma_obs=0.1
H=array(rep(sigma_obs^2*diag(2),n_pre),dim=c(2,2,n_pre))

baseline2_100m<- SDE$new(formulas = formulas,data = dataBE2,type = "RACVM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"),map=list("log_lambda"=factor(c(1,2,rep(NA,4)))))

baseline2_100m$update_lambda(init_lambda)
baseline2_100m$fit()

@@

```{r get results baseline2}

estimates_bas2_25m=as.list(baseline2_25m$tmb_rep(),what="Est")
std_bas2_25m=as.list(baseline2_25m$tmb_rep(),what="Std")

estimates_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Est")
std_bas2_50m=as.list(baseline2_50m$tmb_rep(),what="Std")

estimates_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Est")
std_bas2_75m=as.list(baseline2_75m$tmb_rep(),what="Std")

estimates_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Est")
std_bas2_100m=as.list(baseline2_100m$tmb_rep(),what="Std")

#plot parameters

xmin=list("DistanceShore"=0.07)
xmax=list("DistanceShore"=5)

plots_bas2_25m=baseline2_25m$get_all_plots(baseline=NULL,model_name="baseline2_25m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_50m=baseline2_50m$get_all_plots(baseline=NULL,model_name="baseline2_50m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_75m=baseline2_75m$get_all_plots(baseline=NULL,model_name="baseline2_75m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
plots_bas2_100m=baseline2_100m$get_all_plots(baseline=NULL,model_name="baseline2_100m",xmin=xmin,xmax=xmax,show_CI="pointwise",save=TRUE)
```

```{r estimates on parameter scale baseline2}

post_coeff_bas2_25m=baseline2_25m$post_coeff(n_post=1000)
post_par_bas2_25m=list("tau0"=exp(post_coeff_bas2_25m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_25m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_25m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_25m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_25m$log_lambda[,2])))
mean_par_bas2_25m=lapply(post_par_bas2_25m,mean)
sd_par_bas2_25m=lapply(post_par_bas2_25m,sd)

post_coeff_bas2_50m=baseline2_50m$post_coeff(n_post=1000)
post_par_bas2_50m=list("tau0"=exp(post_coeff_bas2_50m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_50m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_50m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_50m$log_lambda[,2])))
mean_par_bas2_50m=lapply(post_par_bas2_50m,mean)
sd_par_bas2_50m=lapply(post_par_bas2_50m,sd)

post_coeff_bas2_75m=baseline2_75m$post_coeff(n_post=1000)
post_par_bas2_75m=list("tau0"=exp(post_coeff_bas2_75m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_75m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_75m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_75m$log_lambda[,2])))
mean_par_bas2_75m=lapply(post_par_bas2_75m,mean)
sd_par_bas2_75m=lapply(post_par_bas2_75m,sd)

#get 95% confidence intervals
post_coeff_bas2_100m=baseline2_100m$post_coeff(n_post=1000)
post_par_bas2_100m=list("tau0"=exp(post_coeff_bas2_100m$coeff_fe[,"tau.(Intercept)"]),
              "nu0"=exp(post_coeff_bas2_100m$coeff_fe[,"nu.(Intercept)"]),
              "omega0"=post_coeff_bas2_100m$coeff_fe[,"omega.(Intercept)"],
              "sigma_tau"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,1])),
              "sigma_nu"=1/sqrt(exp(post_coeff_bas2_100m$log_lambda[,2])))
mean_par_bas2_100m=lapply(post_par_bas2_100m,mean)
sd_par_bas2_100m=lapply(post_par_bas2_100m,sd)
```

```{r smooth omega baseline2}

>>>>

---
title: "Preliminary analysis of Greenland narwhals trajectories with smoothSDE"
author: "Alexandre Delporte"
date: "2024-06-07"
output:
  pdf_document: default
  html_document: default
  word_document: default
header-includes:
- \usepackage{float}

---

In this analysis, we fit SDE models to the narwhal data. In this preliminary work, the analysis is made without considering any constraints on the motion. Only the effect of the distance to the ship will be considered in the response model.

# Set up

@@

```{r packages,message=FALSE}
set.seed(42)

library(smoothSDE) #sde fit
library(ggplot2)  #plots
library(dplyr)    #manipulate dataframes
library(tidyr)    # same
library(plotly)   # dynamic plots
library(htmlwidgets)  #save dynamic plots
library(mgcv)    # mixed models
library(here)    # path management
library(xtable)  # for latex tables
library(sf)      # read shapefile for land polygons
library(gganimate) # animated trajectories

@@

# Set the path to the directory containing the data
narwhal_data_path <- here("Data", "preprocessed_data","narwhals")

#data before exposure with first 24h removed
dataBE24=read.csv(file.path(narwhal_data_path,"DataBE1.csv"), 
                  header = TRUE,dec = ".")

#data before exposure with first 12h removed
dataBE12=read.csv(file.path(narwhal_data_path,"DataBE2.csv"),
                  header = TRUE,dec = ".")

@@

N=length(unique(dataBE12$ID))

n_pre12=length(dataBE12$time) #nb of obs before exposure with first 12h removed
n_pre24=length(dataBE24$time) #nb of obs before exposure with first 24h removed
n_post=length(dataAE$time) #number of observations after exposure

@@

# Baseline CTCRW models

## CTCRW without random individual effects

We fit a baseline CTCRW with constant parameters (no covariates) on the data before exposure \texttt{dateBE12}. The mean velocity $\mu$ is fixed to $(0,0)$ because we are not investigating a long term migration movement.

```{r fit baseline ctcrw without random effects,message=FALSE}

#initial parameters
par0 <- c(0,0,1,4)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~1,nu=~1)

#Fit baseline with measurement error estimated from the data
baseline_ctcrw0<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                          par0 = par0,
                          other_data=list("log_sigma_obs0"=log(0.05)),
                          fixpar=c("mu1","mu2"))
baseline_ctcrw0$fit()

We now extract the estimated parameters and their standard deviations. These estimates are on the log-scale.

```{r results baseline_ctcrw0}

estimates_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Est")
std_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Std")

We process the results and show a table with the estimates and their standard deviations.

```{r render coefficient estimates table baseline_ctcrw0 }

# Function to extract estimate values into a data frame format
combine_to_df <- function(estimates, std_dev, param_name) {
  data.frame(
    Parameter = param_name,
    Estimate = unlist(estimates, use.names = FALSE),
    Std_Error = unlist(std_dev, use.names = FALSE)
  )
}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_ctcrw0$log_sigma_obs, 
                               std_bas_ctcrw0$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_ctcrw0$coeff_fe, 
                          std_bas_ctcrw0$coeff_fe, 
                          rownames(estimates_bas_ctcrw0$coeff_fe))

# Combine all data frames into one
table_data_baseline_ctcrw0 <- bind_rows(log_sigma_obs, coeff_fe)

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data_baseline_ctcrw0, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
```

From this we can obtain $95\%$ confidence intervals for the parameters $\tau$, $\nu$ and $\sigma_{obs}$.

```{r CI baseline_ctcrw0 parameters}

post_coeff_bas_ctrcw0=baseline_ctcrw0$post_coeff(n_post=10000)
post_par_bas_ctcrw0=
  list("sigma_obs"=1000*exp(post_coeff_bas_ctrcw0$log_sigma_obs),
              "tau"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"tau.(Intercept)"]),
              "nu"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"nu.(Intercept)"]))
mean_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,mean)
sd_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,sd)
quant_par_bas_ctcrw0 <- lapply(post_par_bas_ctcrw0, 
                               quantile, probs = c(0.025, 0.975))

We include the results in a latex table.

```{r CI table baseline_ctcrw0 parameters,results='asis',message=FALSE}

parameter_names <- names(mean_par_bas_ctcrw0)
estimates <- mean_par_bas_ctcrw0
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw0[[param]][1], 
          quant_par_bas_ctcrw0[[param]][2])
})

table_data <- data.frame(
  Parameter = c("$\\sigma_{obs}$","$\\tau$", "$\\nu$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw0),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```{r,results='asis',echo=FALSE}
# Create an xtable object
xtab <- xtable(
  table_data,
  caption = "Baseline estimations",
  label = "table:baseline_estimations"
)

# Customize xtable options
print(
  xtab,
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),
  table.placement="H"
)
```

The estimate for $\tau$ is $\bar \tau =1.08$ h and the estimate for $\nu$ is $\bar \nu =4.69$ km/h. The measurement error standard deviation is estimated to $\bar \sigma_{obs}\simeq 38$ m. The measurement error in Fastloc GPS data mainly depends on the number of satellite that were used to compute the location. This number might change from a data point to another. However, we don't know the number of satellites that computed the position in the data. That is why we assume a constant standard deviation for the measurement error. We know from Weensveen paper, table 3, that it should be at most around 30m when only 4 satellites are involved, and then decrease as the number of satellites increases. This suggest that the measurement error might be slightly overestimated here.

## Sensitivity to measurement error standard deviation

To investigate the sensitivity of the parameters estimates to the measurement error, we fix $\sigma_{obs}$ to different values and estimate the parameters again.

```{r fit baseline ctcrw fixed error without random effects,message=FALSE}

#fixed values for std of measurement error
sigma_obs=c(0.025,0.05,0.075,0.1)
list_baseline_ctcrw0=list()

for (i in 1:length(sigma_obs)){

  H=array(rep(sigma_obs[i]^2*diag(2),n_pre12),dim=c(2,2,n_pre12))

  baseline_ctcrw0_fixed_error<- SDE$new(formulas = formulas,data = dataBE12,
                                        type = "CTCRW",
                              response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))
  baseline_ctcrw0_fixed_error$fit()
  list_baseline_ctcrw0[[i]]= baseline_ctcrw0_fixed_error
}

```{r results for baselines with fixed measurement error}
list_baseline_ctcrw0_est=list()
list_baseline_ctcrw0_std=list()

for (i in seq_along(list_baseline_ctcrw0)) {
  estimates=as.list(list_baseline_ctcrw0[[i]]$tmb_rep(),what="Est")
  std=as.list(list_baseline_ctcrw0[[i]]$tmb_rep(),what="Std")
  list_baseline_ctcrw0_est[[i]]=estimates
  list_baseline_ctcrw0_std[[i]]=std
}

```{r render estimates table baselines with fixed mesaurement error,results='asis', message=FALSE}

for (i in seq_along(list_baseline_ctcrw0)) {
  estimates=list_baseline_ctcrw0_est[[i]]
  std=list_baseline_ctcrw0_std[[i]]

  # Create data frames for each set
  log_sigma_obs <- combine_to_df(estimates$log_sigma_obs, 
                               std$log_sigma_obs, "log_sigma_obs")
  coeff_fe <- combine_to_df(estimates$coeff_fe, 
                          std$coeff_fe, 
                          rownames(estimates$coeff_fe))

  # Combine all data frames into one
  table_data <- bind_rows(log_sigma_obs, coeff_fe)

  latex_table <- xtable(
    table_data, 
    caption = paste(
      "Parameter Estimates and Standard Errors for fixed \\(\\sigma_{obs}=\\) ",
      sigma_obs[i]," m"),
    xtable.table.placement="H")

  # Print LaTeX table
  print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")

}

```{r plot tau and nu estimates as a function of measurement error}
tau_estimates=sapply(list_baseline_ctcrw0_est,
                     function(estimates) 
                       {exp(estimates$coeff_fe["tau.(Intercept)",1])})
nu_estimates=sapply(list_baseline_ctcrw0_est,
                    function(estimates) 
                      {exp(estimates$coeff_fe["nu.(Intercept)",1])})
tau_std=sapply(list_baseline_ctcrw0_std,
               function(std) {exp(std$coeff_fe["tau.(Intercept)",1])})
nu_std=sapply(list_baseline_ctcrw0_std,
              function(std) {exp(std$coeff_fe["nu.(Intercept)",1])})
df=as.data.frame(cbind(tau_estimates,nu_estimates,sigma_obs))
colnames(df)=c("tau","nu","sigma_obs")
df <- df %>%
  pivot_longer(cols = c(tau, nu), names_to = "parameter", values_to = "value")

ggplot(df, aes(x = 1000*sigma_obs, y = value)) +
  geom_point() +
  geom_line() +
  theme_minimal() +
  xlab("Sigma_obs") +
  ylab("Value") +
  ggtitle("Estimated Tau and Nu as functions of fixed Sigma_obs") +
  facet_wrap(~parameter, scales = "free_y")  # Facet by 'parameter' (tau or nu)

We see a clear effect of the fixed measurement error on the estimates of $\tau$, and a slight influence of $\nu$. We also remark that as the measurement error grows, the uncertainty on the estimates of $\tau$ and $\nu$ increases. Overall, knowing or estimating precisely the measurement error will be crucial in the analysis.

### CTCRW with random individual effects

```{r fit baseline ctcrw with random effects,cache=TRUE,message=FALSE}

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"))

#Fit baseline with measurement error estimated from the data
baseline_ctcrw1<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
                    fixpar=c("mu1","mu2"))
baseline_ctcrw1$fit()

```{r results baseline_ctcrw1}

estimates_bas_ctcrw1=as.list(baseline_ctcrw1$tmb_rep(),what="Est")
std_bas_ctcrw1=as.list(baseline_ctcrw1$tmb_rep(),what="Std")

```{r render estimates table baseline_ctcrw1}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_ctcrw1$log_sigma_obs,
                               std_bas_ctcrw1$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_ctcrw1$coeff_fe,
                          std_bas_ctcrw1$coeff_fe,
                          rownames(estimates_bas_ctcrw1$coeff_fe))
log_lambda <- combine_to_df(estimates_bas_ctcrw1$log_lambda, 
                            std_bas_ctcrw1$log_lambda, 
                            rownames(estimates_bas_ctcrw1$log_lambda))
coeff_re <- combine_to_df(estimates_bas_ctcrw1$coeff_re, 
                          std_bas_ctcrw1$coeff_re, 
                          rownames(estimates_bas_ctcrw1$coeff_re))

# Combine all data frames into one
table_data_baseline_ctcrw1 <- bind_rows(log_sigma_obs, coeff_fe, log_lambda, coeff_re)

```{r, results='asis', echo=FALSE}
latex_table <- xtable(table_data_baseline_ctcrw1, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
```

As previously, we can obtain $95\%$ confidence intervals for the parameters $\tau$, $\nu$ and $\sigma_{obs}$, and include the results in a latex table.

```{r CI baseline_ctcrw1 parameters}

post_coeff_bas_ctrcw1=baseline_ctcrw1$post_coeff(n_post=10000)
post_par_bas_ctcrw1=list(
  "sigma_obs"=1000*exp(post_coeff_bas_ctrcw1$log_sigma_obs),
  "tau"=exp(post_coeff_bas_ctrcw1$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_ctrcw1$coeff_fe[,"nu.(Intercept)"]),
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_ctrcw1$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_ctrcw1$log_lambda[,"nu.s(ID)"])))

mean_par_bas_ctcrw1=lapply(post_par_bas_ctcrw1,mean)
sd_par_bas_ctcrw1=lapply(post_par_bas_ctcrw1,sd)
quant_par_bas_ctcrw1 <- lapply(post_par_bas_ctcrw1, 
                               quantile, probs = c(0.025, 0.975))

```{r CI table baseline_ctcrw1 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_ctcrw1)
estimates <- mean_par_bas_ctcrw1
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw1[[param]][1], 
          quant_par_bas_ctcrw1[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\sigma_{obs}$","$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw1),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```
```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "Baseline estimations with random effects",
  label = "table:baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```

Note that the random effect is multiplicative and not additive since it appears in the $\log$ of the parameters. We can plot the estimates on the parameter scale for each narwhal.

```{r plot baseline_ctcrw1 estimates,results='hide'}

baseline_ctcrw1$get_all_plots(baseline=NULL,show_CI="simultaneous")

### RACVM without random individual effects

We can also test whether adding a constant rotational $\omega$ in the SDE model would make sense. For this, we fit a RACVM model.

```{r fit baseline racvm,message=FALSE}

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~1,
                 nu=~1,
                 omega=~1)

#Fit baseline with measurement error estimated from the data
baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
                          response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
                    fixpar=c("mu1","mu2"))
baseline_racvm0$fit()

@@

```{r results baseline_racvm0}

estimates_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Est")
std_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Std")

@@

```{r render estimates table baseline_racvm0}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_racvm0$log_sigma_obs, 
                               std_bas_racvm0$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_racvm0$coeff_fe,
                          std_bas_racvm0$coeff_fe, 
                          rownames(estimates_bas_racvm0$coeff_fe))

# Combine all data frames into one
table_data_baseline_racvm0 <- bind_rows(log_sigma_obs, coeff_fe)

@@

```{r, results='asis', echo=FALSE}
latex_table <- xtable(table_data_baseline_racvm0, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
```

The estimate of $\omega$ is $0$. This indicates that a \textbf{constant} parameter $\omega$ is irrelevant in the model. As a consequence, the final baseline model we keep is \texttt{baseline\_ctcrw1} which corresponds to a CTCRW with constant parameters $\tau$ and $\nu$ and mean velocity fixed to zero.

## Sensitivity analysis of the baseline CTCRW to the tagging effect threshold

In this first analysis, we only removed 12h of data after the tagging for each narwhal. However, the narwhals might still be affected by the tagging and not behave "normally". This would invalidate all our conclusions in the response model, if the baseline model happens to be biased. Therefore, we need to assess the sensitivity of the baseline model to the time threshold we choose for the tagging effect. Here, we only repeat the fit for the model \texttt{baseline\_ctcrw1} with the sub dataset \texttt{dataBE24} where all the data until 24h after tagging have been removed.

```{r fit baseline sensitivity analysis tagging effect,message=FALSE}

#initial parameters
par0 <- c(0,0,1,4)

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"))

#Fit baseline with measurement error estimated from the data
baseline_ctcrw1_24h<- SDE$new(
  formulas = formulas,data = dataBE24,type = "CTCRW",
  response = c("x","y"),par0 = par0,
  other_data=list("log_sigma_obs0"=log(0.05)),
  fixpar=c("mu1","mu2"))

baseline_ctcrw1_24h$fit()

@@

```{r results baseline_ctcrw1_24h}

estimates_bas_ctcrw1_24h=as.list(baseline_ctcrw1_24h$tmb_rep(),what="Est")
std_bas_ctcrw1_24h=as.list(baseline_ctcrw1_24h$tmb_rep(),what="Std")

@@

```{r render estimates table baseline_ctcrw1_24h}

# Create data frames for each set
log_sigma_obs <- combine_to_df(
  estimates_bas_ctcrw1_24h$log_sigma_obs, 
  std_bas_ctcrw1_24h$log_sigma_obs, "log_sigma_obs")

coeff_fe <- combine_to_df(estimates_bas_ctcrw1_24h$coeff_fe, 
                          std_bas_ctcrw1_24h$coeff_fe, 
                          rownames(estimates_bas_ctcrw1_24h$coeff_fe))

# Combine all data frames into one
table_data_baseline_ctcrw1_24h <- bind_rows(log_sigma_obs, coeff_fe)
